---
title: "Showing Averages Within Large Datsets"
description: |
  Using the geomborderline package to make averages jump out.
author:
  - name: Nick Zani
    url: {}
date: 2022-01-25
output:
  distill::distill_article:
    self_contained: false
---

### Introduction

Last week I used the [ggborderline package](https://wurli.github.io/ggborderline/) to tidy up line graphs. This week I used the same data again to try and show how the average has changed over time, while still showing the overall distribution.

### Data

I chose 15 states from the Tidy Tuesday data on the number of nurses employed in each US state by month. I then created three more data frames. One showing the max and min values by year, one showing the average, and one for the final labels.

```{r eval=TRUE}
library(tibble)
library(dplyr)
library(stringr)
library(readr)
library(tidyr)
library(janitor)
library(ggplot2)

nurses <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-10-05/nurses.csv') %>%
  clean_names()

states <- nurses %>%
  filter(year == min(year)) %>%
  arrange(total_employed_rn) %>%
  head(30) %>%
  tail(15) %>%
  pull(state)

nurses_filtered <- nurses %>%
  filter(state %in% states)

max_min <- nurses_filtered %>%
  group_by(year) %>%
  summarise(min_employed = min(total_employed_rn, na.rm = TRUE),
            max_employed = max(total_employed_rn, na.rm = TRUE))

averages <- nurses_filtered %>%
  group_by(year) %>%
  summarise(mean_employed = mean(total_employed_rn, na.rm = TRUE),
            median_employed = median(total_employed_rn, na.rm = TRUE))

labels <- averages %>%
  ungroup() %>%
  filter(year == max(year)) %>%
  mutate(label = "Median")

```

### Building Up the Plot

The plot is made up of four parts:
- The ribbon between the maximum and minimum points, with the colour defined in the fill option
- The line of each individual state
- The average line, passed to geom_borderline so it jumps out against the other lines
- The secondary axis for labelling

```{r fig.width = 8, fig.height = 6}

library(ggborderline)

ggplot(max_min, aes(year)) +
  geom_ribbon(aes(ymin = min_employed - 200, ymax = max_employed + 200), fill = "snow2") +
  geom_line(data = nurses_filtered, aes(x = year, y = total_employed_rn, group = state), colour = "grey 75", size = 1) +
  geom_borderline(data = averages, aes(x = year, y = median_employed), size = 2, bordersize = 1.5, colour = "royalblue4") +
  theme_classic() +
  scale_y_continuous(labels = scales::comma_format(),
                     sec.axis = dup_axis(breaks = labels$median_employed,
                                         labels = labels$label,
                                         name = "")) +
  scale_x_continuous(expand = expansion(mult = c(0.05, 0.01))) +
  labs(title = "Nurse Numbers in Selected States", 
       subtitle = "geom_ribbon + geom_borderline\n",
       x = "",
       y = "Count of Nurses\n",
       caption = "\nNick Zani | @nickzani") +
  theme(strip.text = element_text(size = 18),
        axis.title = element_text(size = 14, colour = "grey 15"),
        axis.text.x = element_text(size = 18),
        axis.text.y = element_text(size = 18),
        plot.title = element_text(size = 18, colour = "grey 15"),
        plot.subtitle = element_text(size = 14, colour = "grey 15"),
        plot.caption = element_text(size = 12, colour = "grey 15"),
        legend.text = element_text(size = 16),
        legend.title = element_text(size = 18),
        legend.position = "none",
        axis.line.y.right = element_line(color = "white"),
        axis.ticks.y.right = element_line(color = "white"),
        axis.text.y.right = element_text(color = "royalblue4")
  )

```