---
title: "Tidying up Line Graphs"
description: |
  Using ggborderline and secondary axis labels.
author:
  - name: Nick Zani
    url: {}
date: 2022-01-17
output:
  distill::distill_article:
    self_contained: false
---

### Introduction

Last week I had a play around with the [geomtextpath](https://github.com/AllanCameron/geomtextpath) package, and this week I wanted to expand on that with [ggborderline](https://wurli.github.io/ggborderline/), again using some Tidy Tuesday data from October 2021 on the number of nurse numbers in different American states.

### Data
No data manipulation to be done here, just import and filter the data for some selected states. I chose states which had similar numbers so that some of the lines in the plots would cross over later.

```{r eval=TRUE}
library(tibble)
library(dplyr)
library(stringr)
library(readr)
library(tidyr)
library(janitor)
library(ggplot2)

nurses <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-10-05/nurses.csv')


nurses_filtered <- nurses %>%
  clean_names() %>%
  filter(state %in% c("Alabama", "Arizona", "Oregon", "Iowa"))

```

### Initial Visualisation

The off-the-shelf ggplot looks fine, but there are definitely some improvements we can make.  

```{r fig.width = 7, fig.height = 6}

nurses_filtered %>%
  ggplot(aes(x = year, y = total_employed_rn, group = state, colour = state)) +
  geom_line(size = 2) +
  scale_colour_brewer(type = "qual", palette = 2) +
  theme_classic() +
  scale_y_continuous(labels = scales::comma_format()) +
  labs(title = "Nurse Numbers in Selected States", 
       subtitle = "Standard ggplot\n",
       colour = "State",
       x = "",
       y = "Count of Nurses\n",
       caption = "\nNick Zani | @nickzani") +
  theme(strip.text = element_text(size = 18),
        axis.title = element_text(size = 14),
        axis.text.x = element_text(size = 14),
        axis.text.y = element_text(size = 18),
        plot.title = element_text(size = 18),
        plot.subtitle = element_text(size = 14),
        plot.caption = element_text(size = 12),
        legend.text = element_text(size = 16),
        legend.title = element_text(size = 18)
  )

```

### Using ggborderline

The [ggborderline](https://wurli.github.io/ggborderline/) is a great package for tidying up line graphs where the lines overlap. It adds in a white border to the lines which can be tweaked in thickness

```{r fig.width = 7, fig.height = 6}

library(ggborderline)

nurses_filtered %>%
  ggplot(aes(x = year, y = total_employed_rn, group = state, colour = state)) +
  geom_borderline(size = 2, bordersize = 0.75) +
  scale_colour_brewer(type = "qual", palette = 2) +
  theme_classic() +
  scale_y_continuous(labels = scales::comma_format()) +
  labs(title = "Nurse Numbers in Selected States", 
       subtitle = "With ggborderline\n",
       colour = "State",
       x = "",
       y = "Count of Nurses\n",
       caption = "\nNick Zani | @nickzani") +
  theme(strip.text = element_text(size = 18),
        axis.title = element_text(size = 14),
        axis.text.x = element_text(size = 14),
        axis.text.y = element_text(size = 18),
        plot.title = element_text(size = 18),
        plot.subtitle = element_text(size = 14),
        plot.caption = element_text(size = 12),
        legend.text = element_text(size = 16),
        legend.title = element_text(size = 18)
  )

```

### Improving the Labels

The final tweak is to fix the labels. The legend is not in order, and makes the graph harder to read. By adding a secondary axis in, we can simplify this. The trick is to create a second dataframe which contains just the labels at the maximum values. 

```{r fig.width = 7, fig.height = 6, preview=TRUE}

library(ggborderline)

max_y_value <- nurses_filtered %>%
  filter(year == max(year))

nurses_filtered %>%
  ggplot(aes(x = year, 
             y = total_employed_rn, 
             group = state, 
             colour = state)) +
  geom_borderline(size = 2, bordersize = 0.75) +
  scale_colour_brewer(type = "qual", palette = 2) +
  theme_classic() +
  scale_y_continuous(labels = scales::comma_format(),
                     sec.axis = dup_axis(breaks = max_y_value$total_employed_rn,
                                         labels = max_y_value$state,
                                         name = "")) +
  labs(title = "Nurse Numbers in Selected States", 
       subtitle = "With ggborderline + secondary axis labelling\n",
       colour = "State",
       x = "",
       y = "Count of Nurses\n",
       caption = "\nNick Zani | @nickzani") +
  theme(strip.text = element_text(size = 18),
        axis.title = element_text(size = 14),
        axis.text.x = element_text(size = 14),
        axis.text.y = element_text(size = 18),
        plot.title = element_text(size = 18),
        plot.subtitle = element_text(size = 14),
        plot.caption = element_text(size = 12),
        legend.text = element_text(size = 16),
        legend.title = element_text(size = 18),
        legend.position = "none",
        axis.line.y.right = element_line(color = "white"),
        axis.ticks.y.right = element_line(color = "white")
  )

```